Patterns of stress tolerance versus resistance in the thermal reactionomes from populations of cold and warm-climate ant species

### 2014-08-01 doc/ApTranscriptome_manuscript.Rmd
Revised Intro incorporating comments from AE/NG

---
title: "Patterns of stress tolerance versus resistance in genome-wide expression data
  of parapatric ant species"
author: \centering{John Stanton-Geddes\textsuperscript{1}, Andrew D. Nguyen\textsuperscript{1}, Mahesh Vangala\textsuperscript{2}, James Vincent\textsuperscript{2}, Robert R. Dunn\textsuperscript{3}, Nathan Sanders\textsuperscript{4}, Nicholas J. Gotelli\textsuperscript{1}, Sara Helms Cahan\textsuperscript{1} \\ \textsuperscript{1}: Department of Biology, University of Vermont \\ \textsuperscript{2}: Vermont Genetics Network, University of Vermont \\ \textsuperscript{3}: North Carolina State University \\ \textsuperscript{4}: University of Copenhagen}
csl: ecology.csl
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  html_document: null
  word_document: default
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

Thermal tolerance is a key trait governing the distributions of species and underlies their future responses to climate change. However, the full set of genes and pathways underlying thermal tolerance are poorly understood for most species. Here, we demonstrate how insights about the genetic basis of local adaption to climate can be gained by applying a reaction norm approach to transcriptomic data. We introduce the thermal *reactionome*, the characterization of the thermal reaction norm of all genes in an organism's transcriptome. We characterized transcriptome-wide gene expression over a 38-degree thermal range for two closely-related ant species, *A. carolinensis* and *A. picea*, which are parapatrically distributed along a temperature gradient in eastern deciduous forests of North America. At least 10% of genes are thermally responsive. When the two species were compared, the warm-adapted *A. carolinensis* displayed higher basal levels of gene expression, \todo{a lower magnitude of upregulation at thermal extremes}, upregulated of fewer heat shock proteins, and greater inertia of changes in gene expression in response to thermal extremes. This pattern is consistent with both constitutive thermal defenses and enhanced proteome stability. Moreover, a surprisingly large proportion of genes showed divergent patterns of response to temperature between the two species, suggesting that they have different strategies for coping with temperature stress. *Aphaenogaster carolinensis* shut down more genes than *A. picea* in response to thermal extremes, whereas the same genes tended to be upregulated in *A. picea*. These included genes functionally involved in \todo{metabolism, cellular organization and repair}. Altogether, these results suggest that *A. picea* actively resists thermal shock by synthesizing heat shock proteins and other protective molecules to maintain function, whereas *A. carolinensis* slows down metabolism and relies upon other mechanisms to tolerate thermal stress. More broadly, this study adds to a growing body of evidence that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention, despite enzyme efficiency tradeoffs that such mechanisms entail.

@@

Temperature plays a prominent role in structuring biological diversity from molecular to macroecological scales [@kingsolver2009]. The ability of individuals to maintain function and survive thermal extremes, thermal tolerance, varies considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013]. These differences in thermal tolerance are largely genetic [@krebs1997a; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011]. How this variation of thermal tolerance evolved, and its potential for future evolution [@hoffman2008], is not well understood. To address this, we need information on the extent to which geographic differentiation and selection has acted upon the diversity of genes, and their expression patterns, involved in thermal tolerance.

The study of whole organism gene expression (e.g. transcriptomics) has increased in tandem with improved RNA sequencing (RNAseq) technology. In evolutionary ecology, RNAseq studies to date have largely focused on differential expression across few environments [@meyer2011; @teets2012; @pespeni2013; @barshis2013; @fu2014] or tissues [@li2013] using an ANOVA-type statistical approach. However, many ecologically-relevant aspects of the environment such as temperature vary continuously in nature. Variation in the phenotype of an organism in response to such a continuous predictor (i.e. phenotypic plasticity) is referred to as a reaction norm. Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014] and differences in the shape of reaction norms are often greater than differences in mean values of either species or populations [@murren2014]. Murren et al [-@murren2014] concluded that many patterns of biologically-relevant plasticity are likely to be missed in studies with few environments. Similarly, gene expression studies that look at just a few environments (e.g. hot vs. cold) are likely to miss key differences among species that are due not just to whether it is hot, but rather how hot it is.

Moreover, the results provided by RNAseq data do not always directly address the aims of studies. For evolutionary ecology studies, the goal of RNAseq studies often is to identify the genetic basis of environmental adaptation [@meyer2011; @teets2012; @pespeni2013; @barshis2013; @oneil2014]. However, lists of differentially-expressed genes derived from RNAseq results are prone to false positives [@gonzalez2013], miss genes with small expression changes that have large phenotypic effects [@padovan2013] and may not be directly informative about organismal status [@feder2005]. Genome-wide association (GWA) studies have faced a similar problem: phenotypic effects of the few statistically-significant genes are over-estimated while many causal genes of small average effect are missed [@beavis1994; @larsson2013]. Both our understanding of the genetic basis of traits, and our ability to predict evolutionary potential, are undermined by these issues. Thus, plant and animal breeders have largely abandoned GWA in favor of genomic prediction [@goddard2007; @yang2010].

Here, we present a novel method for RNAseq analysis that characterizes the reaction norm for all genes in an organism's transcriptome across a thermal gradient; the *thermal reactionome*. Instead of focusing on individual genes with differential expression, we compare overall patterns in such reactionomes between closely-related species to gain insight into how spatially divergent selection has changed patterns of gene expression. We apply this approach in a comparative study of two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996], which have more southern and northern distributions, respectively, and diverged approximately .... years ago [bernice??].

Temperature is a major selective pressure as *Aphaenogaster* colonies experience freezing winters and hot summers throughout the year, and at shorter time scales, individual workers can experience large thermal changes as they leave the buffered ant nest to forage for food. *A. carolinensis* lives in a climate that is hotter and has a longer growing season than that of *A. picea* [@diamond2013]. Thus, patterns of gene expression underlying tolerance of thermal extremes is likely to differ between these species in a number of ways.

First, the number of genes showing plasticity of expression to thermal extremes may differ between the species. Finding more genes are up-regulated to high temperatures in the colder-climate species *A. picea* would indicate that this species relies on activation of the cellular stress response [@...] and other protective genes to a greater extent than *A. carolinensis* , while the opposite would be true for cold temperatures. Conversely, finding that more genes are down-regulated in response to thermal extremes would indicate that a species relies upon proteome stability [@...] and behavioral mechanisms [@sunday2014] to tolerate thermal stress.

Second, responsive genes may change their sensitivity to thermal extremes. Finding that heat-responsive genes are up-regulated at higher temperatures in *A. carolinensis* than *A. picea* would be consistent with local adaptation for greater heat tolerance. That is, more extreme temperatures are required for activation of the heat shock response [@somero...]. Conversely, finding cold-responsive genes are up-regulated at lower temperatures in *A. picea* than *A. carolinensis* would indicate local adaptation for cold tolerance. Similar patterns may be observed for genes that are down-regulated in response to thermal extremes.

Moreover, insight on the evolution of thermal tolerance can be gained by examining shifts in the patterns of responsive genes between species. For example, if adaptation to thermal extremes works by genetic assimilation [@pigliucci2006], genes up-regulated in response to cold temperatures in *A. carolinensis* may have increased basal expression in the colder-climate species *A. picea*.

Our overall goal in this study was to determine the extent to which geographic differentiation and selection has changed patterns of gene expression involved in thermal tolerance in these species. Specifically, we (i) generated the first transcriptomic data for the diverse and abundant temperate ant genus *Aphaenogaster* in eastern North America, (ii) identified thermally-responsive genes in each species, (iii) compared the temperatures of gene activation and sensitivity of the thermally-responsive genes in each species, (iv) looked for expression patterns between the two species consistent with genetic assimilation, (v) evaluated the extent to which changes in reaction norms were due to changes in the mean or shape, and (vi) examined the gene ontology terms associated with thermal responsiveness in each species.

# Methods

@@

>>>>

---
title: "Patterns of stress tolerance versus resistance in the thermal reactionomes from populations of cold and warm-climate ant species"
author: \centering{John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1}, James Vincent \textsuperscript{2}, Mahesh Vangala \textsuperscript{2}, Robert R. Dunn \textsuperscript{3}, Nathan J. Sanders \textsuperscript{4}, Nicholas J. Gotelli \textsuperscript{1}, Sara Helms Cahan \textsuperscript{1} \\ \vspace{5 mm} \textsuperscript{1} Department of Biology, University of Vermont \\ \textsuperscript{2} Vermont Genetics Network, University of Vermont \\ \textsuperscript{3} Department of Biological Sciences, North Carolina State University \\ \textsuperscript{4} Center for Macroecology, Evolution and Climate, University of Copenhagen}
csl: ecology.csl
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  html_document: null
  word_document: default
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

**250 words allowed - need to cut 56**

Thermal tolerance is a key trait governing the distributions of species and underlies their future responses to climate change. In this study, we demonstrate how insight on the evolution of thermal tolerance can be gained through characterizing an organism’s thermal “reactionome” – the reaction norm for all genes in an organism's transcriptome measured across a thermal gradient. We characterized the reactionomes of two populations of parapatric ant species of eastern U.S. deciduous forests: the northern cold-climate Aphaenogaster rudis and the southern warm-climate Aphaenogaster carolinensis . After exposing laboratory-acclimated individuals to a range of realistic temperatures (0° - 38.5°C), we found that 10% of genes were thermally responsive. When the two species were compared, the warm-adapted *A. carolinensis* displayed higher basal levels of gene expression, \todo{check this statement} a lower magnitude of upregulation at thermal extremes, upregulated of fewer heat shock proteins, and greater inertia of changes in gene expression in response to thermal extremes. This pattern is consistent with both constitutive thermal defenses and enhanced proteome stability. Moreover, a surprisingly large proportion of genes showed divergent patterns of response to temperature between the two species, suggesting that they have different strategies for coping with temperature stress. *Aphaenogaster carolinensis* shut down more genes than *A. picea* in response to thermal extremes, whereas the same genes tended to be upregulated in *A. picea*. These included genes functionally involved in \todo{check} metabolism, cellular organization and repair. Altogether, these results suggest that *A. picea* actively resists thermal shock by synthesizing heat shock proteins and other protective molecules to maintain function, whereas *A. carolinensis* slows down metabolism and relies upon other mechanisms to tolerate thermal stress. More broadly, this study adds to a growing body of evidence that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention, despite enzyme efficiency tradeoffs that such mechanisms entail.

# Significance Statement

**120-word-maximum statement about the significance of their research paper written at a level understandable to an undergraduate-educated scientist outside their field of specialty**

@@

Temperature plays a prominent role in structuring biological diversity from molecular to macroecological scales [@kingsolver2009]. The ability of individuals to maintain function and survive thermal extremes, thermal tolerance, varies considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013]. While organisms differ in fixed traits such as proteome stability [@ghosh2010] and behavior [@sunday2014] that allow them to function in the typical range of temperatures to which they are exposed, tolerance to thermal extremes often involves rapid changes in gene expression [@kultz2005; @somero2011]. These differences within and among species in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011]. How this variation of thermal tolerance evolved, and its potential for future evolution [@hoffmann2008], is not well understood. To address this, we need information on the extent to which geographic differentiation and selection has acted upon the diversity of genes, and their plasticity of expression, involved in thermal tolerance.

In recent years, whole organism gene expression approaches (e.g. transcriptomics) using high-throughput RNA sequencing (RNAseq) technology have been widely applied to identify genes underlying thermal tolerance [@meyer2011; @teets2012; @barshis2013; @oneil2014], drought tolerance [@villar2011] and many other traits. These RNAseq studies have used an ANOVA-type experimental or sampleing design with only a few levels of temperature or other environmental stressors tested. However, many ecologically-relevant aspects of the environment such as temperature vary continuously in nature. The reaction norm describes variation in the phenotype of a single genotype in response to a continuous environmental gradient [@gomulkiewicz1992]. Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014], and differences in the shape of reaction norms are often greater than differences in mean values of either species or populations [@murren2014]. Murren et al [-@murren2014] concluded that many patterns of biologically-relevant plasticity are likely to be missed with ANOVA-style designs that compare only a few discrete levels of an environmental variable. Similarly, gene expression studies that look at just a few environments (e.g. hot vs. cold) are likely to miss key differences among species that are due not just to whether it is hot, but rather how hot it is.

Moreover, the results provided by RNAseq data do not always directly address the aims of studies. For evolutionary ecology studies, the goal of RNAseq studies often is to identify the genetic basis of environmental adaptation [@meyer2011; @teets2012; @barshis2013; @oneil2014]. However, lists of differentially-expressed genes derived from RNAseq results are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011], may miss genes with small expression changes that have detectable phenotypic effects [@yan2002; @padovan2013], identify genes that are correlated, but not causative, of phenotypes [@fay2004] and may not be directly informative about organismal status [@feder2005]. Genome-wide association (GWA) studies have faced a similar problem: phenotypic effects of a small number of statistically-significant genes are over-estimated while small effects of a large number of other causal genes are missed [@beavis1994; @larsson2013]. Our understanding of the genetic basis of traits, and our ability to predict evolutionary potential, are compromised by these issues. Thus, plant and animal breeders have largely abandoned GWA in favor of genomic prediction [@goddard2007; @yang2010].

Here, we present a novel method for RNAseq analysis: the *thermal reactionome*, which is a characterization of the reaction norm for all genes in an organism's transcriptome across a thermal gradient. Rather than focusing on the identity of individual genes with differential expression, we compare overall patterns in the reactionomes of two closely-related temperate-zone species to gain insight into how spatially divergent selection has changed patterns of thermally-responsive gene expression. We apply this approach in a comparative study of two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996], which have more southern and northern distributions, respectively, and diverged approximately .... years ago (bernice??). Temperature is a potentially strong selective force on ground-nesting *Aphaenogaster* colonies, which must tolerate seasonally freezing winters and hot summers. On shorter time scales, individual workers can experience extreme thermal environments when they leave the thermally buffered ant nest to forage for food [@lubertazzi2012].

*A. carolinensis* lives in a climate that is hotter and has a longer growing season than that of *A. picea* (Supplemental Information). Thus, patterns of gene expression underlying tolerance of thermal extremes is likely to differ between these species in a number of ways. First, the number of genes showing plasticity of expression to thermal extremes may differ between the species. Finding more genes are up-regulated to high temperatures in the colder-climate species *A. picea* would indicate that this species relies on activation of the cellular stress response [@kultz2005] and other protective genes to a greater extent than *A. carolinensis*, while the opposite would be true for cold temperatures. Conversely, finding that more genes are down-regulated in response to thermal extremes would indicate that a species relies to a greater extent upon proteome stability [@ghosh2010] and behavioral mechanisms [@sunday2014] to tolerate thermal stress.

In addition, responsive genes may change in their sensitivity to thermal extremes. Finding that heat-responsive genes are up-regulated at higher temperatures in *A. carolinensis* than *A. picea* would be consistent with local adaptation for greater heat tolerance. That is, more extreme temperatures are required for activation of the heat shock response [@somero2011]. Conversely, finding cold-responsive genes are up-regulated at lower temperatures in *A. picea* than *A. carolinensis* would indicate local adaptation for cold tolerance. Similar patterns may be observed for genes that are down-regulated in response to thermal extremes. Finally, changes in the expression patterns of responsive genes between species can provide insight on the evolution of thermal tolerance. Specifically, if adaptation to thermal extremes works by genetic assimilation [@pigliucci2006], genes up-regulated in response to cold temperatures in *A. carolinensis* may have increased basal expression in the colder-climate species *A. picea*, and vice versa.

Importantly, none of the above hypotheses depend on the functional annotation of the responsive genes, which can be inaccurate in non-model systems [@primmer2013; @skunca2012]. However, insight on the molecular toolkit involved in adaptation to thermal extremes can be gained through circumspect examination of the functional annotation of responsive genes. For example, we expect genes in the cellular stress response [@kultz2005] to be up-regulated in response to thermal extremes, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated in response to thermal extremes [@somero2011].

@@

# Material and Methods

# Figure Legends
